---
title: "Event detection"
author: "Beni Stocker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Event detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(dplyr)
library(ggplot2)
library(GECOr)
library(here)
```

An event is defined here as consecutive time steps where a certain condition is fulfilled - yielding a vector of a logical (boolean) variable. This vignette demonstrates how such events are derived from the boolean vector using the GECOr function `get_consecutive()`, and a time series data frame re-arranged and labelled with sequential time steps into a given event (`eventstep`), using the GECOr function `align_events()`. 

## Prepare data

Read data from the file contained in this repository. This is a data frame containing time series of multiple variables arranged in columns and each time step (day) is a row.
```{r}
df <- readRDS(file = here::here("data/df_ch-lae.rds"))
```

Below, let's consider the following variables:

- `TIMESTAMP`: the date of observation
- `TA_F_MDS`: air temperature in degrees Celsius

## Event detection

Let's consider frost events as periods of consecutive days where air temperature is below zero and create a new variable in the data frame that contains this information as a boolean vector (`isfrost`).

```{r}
df <- df |> 
  mutate(isfrost = ifelse(TA_F_MDS < 0, TRUE, FALSE))
```

With this vector, we can apply the function `get_consecutive()`. 

```{r}
events <- get_consecutive(
  df$isfrost,
  merge_threshold = 3,
  leng_threshold = 5,
  do_merge = TRUE
)
```

This yields information about the events start and length, expressed in index position of the boolean vector (here `df$isfrost`).
```{r}
head(events)
```

We can use this to get the corresponding dates.

```{r}
# get start and end date of longest sequences
events <- events |> 
  mutate(
    start = lubridate::as_date(df$TIMESTAMP[events$idx_start]),
    end = lubridate::as_date(df$TIMESTAMP[events$idx_start + events$len - 1])
  )
```

With this information, we can visualise the detected frost events.

```{r}
ggplot() +
  # grey rectangles for each event
  geom_rect(
    data = events,
    aes(
      xmin = start,
      xmax = end,
      ymin = min(df$TA_F_MDS, na.rm = TRUE),
      ymax = max(df$TA_F_MDS, na.rm = TRUE)
    ),
    fill = "grey80"
  ) +
  geom_line(
    data = df,
    aes(
      TIMESTAMP,
      TA_F_MDS
    )) +
  labs(x = "Time",
       y = "Air temperature (degrees Celsius)") +
  theme_classic()
```

## Align events

Now that the events are detected, the data within each event can be re-arranged, treating each event as a comparable sequence of time steps (here days).

```{r}
aligned <- align_events(
  df, 
  events = events,
  leng_threshold = 5,
  before = 3, 
  after = 100
)
```

We can now plot data during each frost event versus the `idx_event` (here day into frost event, counting from 0, in the output data frame).

```{r}
aligned |> 
  ggplot(aes(idx_event, TA_F_MDS)) +
  geom_point() + 
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_classic()
```


